\documentclass[12pt,a4paper]{article}

% Encoding
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% Page layout
\usepackage{geometry}
\geometry{a4paper,left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm}

% Font - Arial (LSE requirement)
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}

% Spacing
\usepackage{setspace}

% Bibliography
\usepackage[
backend=biber,
style=apa,
url = false,
doi = false,
isbn=false,
]{biblatex}
\addbibresource{bibliography.bib}
\usepackage[hidelinks]{hyperref}

% Headers
\usepackage{fancyhdr}
\setlength{\headheight}{14.5pt}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage}
\renewcommand{\headrulewidth}{0pt}

% Fix hbox warning
\sloppy

% Misc
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{appendix}

\begin{document}

\begin{doublespace}

% Cover page
\thispagestyle{empty}
\pagenumbering{gobble}
\begin{center}
    \vspace*{2cm}
    {\large GV499 Dissertation\\
    MSc Public Policy and Administration \\
    Department of Government\\
    London School of Economics}\\
    \vspace{2cm}
    {\Large\textbf{Assumptions Matter: Central Bank Independence, Epistemology, and Structural Regulatory Capture}}\\
    \vspace{2cm}
    {\large \textit{Dissertation Outline}}\\
    \vspace{0.5cm}
    \today
\end{center}
\newpage

\pagenumbering{arabic}
\setcounter{page}{1}

% -------------------------------------------------------
\section*{Research Question}
Why do private financial interests defend Federal Reserve independence?

\section*{Thesis}
The interests of financial institutions were encoded in the institutional structures governing the Federal Reserve. These interests can be traced from the epistemological assumptions rooted in economics literature on central bank independence, through personnel and discourse, into consistent blockage of legislative revision of the Fed's dual mandate. The end result is a system that reflects regulatory capture without requiring explicit coordination or lobbying among firms; private interests instead only have to promote inaction.

\newpage
\tableofcontents
\newpage

% -------------------------------------------------------
\section{Introduction (\textasciitilde750 words)}

\begin{itemize}
    \item \textbf{HOOK: JPMorgan/Powell paradox} --- JPMorgan defends Powell against a government it otherwise influences.

    \item \textbf{The asymmetry} --- Financial institutions challenge the Fed when it imposes constraints on them (stress tests, capital requirements) but defend it against democratic oversight. Regulatory capture theory would posit the opposite --- that financial institutions can use their power to lobby politicians. Extension: Congress has control over the goals of the Federal Reserve. Why don't they exercise their powers to reform the goals of monetary policy?

    \item \textbf{RQ + thesis preview}

    \item \textbf{Epistemic capture as central contribution} --- The economics literature that defends the merits of CBI bases both modelling and empirical analysis on normative assumptions that mimic the interests of financial firms.

    \item \textbf{``Ingrained'' framing} --- Financial interests weren't subsequently captured by the Fed. They were built in from the start.
\end{itemize}

% -------------------------------------------------------
\section{Background and Theory (\textasciitilde1,900 words)}

\textit{Most of this will be further developed in Section 4.1.}

\subsection{The Economics Literature on CBI (\textasciitilde700 words)}

\subsubsection{Central banking and CBI defined}

\begin{itemize}
    \item \textbf{Goal vs.\ instrument independence (Debelle and Fischer 1994)} --- Congress sets the goals; the Fed has operational independence to achieve those goals. Powell himself says this. This distinction matters for the analysis of CBI indices, with the price stability component being incorporated into CBI scores.

    \item \textbf{Fed's dual mandate (Section 2A, Federal Reserve Act)} --- Simple definition. Worth noting this is actually a ``tripartite'' mandate.
\end{itemize}

\subsubsection{Theoretical foundations}

\begin{itemize}
    \item \textbf{Kydland and Prescott (1977)} --- Time-inconsistency problem applied to the inflation-unemployment tradeoff. The general principle for CBI.

    \item \textbf{Barro and Gordon (1983)} --- Applies the logic directly to monetary authorities. Monetary policy is neutral in the long run, so it can only reap benefits through short-run surprise. Under discretion, this produces an inflationary bias.

    \item \textbf{Rogoff (1985)} --- The conservative central banker solution: appoint a monetary authority who prefers less inflation than society does. Explicitly third-best --- behind removing the underlying labour market distortion (first-best, itself a normative claim about labour markets) and a constitutional rule (second-best). Conservative central bankers trade lower employment for lower inflation. The empirical literature inherits the conclusion but drops all the caveats.

    \item \textbf{Posen (1995)} --- The conservative central banker's preferences are mirrored by those of the financial sector; it is the financial sector's support that determines levels of CBI.
\end{itemize}

\subsubsection{Empirical development}

\begin{itemize}
    \item \textbf{GMT (1991); CWN (1992); Alesina and Summers (1993)} --- Observable institutional characteristics compiled into indices; statistical models measuring how CBI influences output and inflation. The free lunch claim: lower inflation, no output cost. Alesina and Summers themselves caveat this as ``not conclusive'' --- that caveat disappears from the policy consensus.

    \item \textbf{Romelli (2022); Adrian et al.\ (2024)} --- Same general approach, updated and extended. Romelli's objectives dimension --- which encodes price stability conservatism --- has more than doubled across his sample period. He presents this approvingly.

    \item \textbf{Shared features across all five} --- De jure measurement, inherited time-inconsistency prior, price stability as the benchmark, free lunch claim that drops Rogoff's acknowledged trade-off. They don't argue against employment as an objective --- they assume it away by construction.

    \item \textbf{Isabel Schnabel on monetary dominance} --- Notes that the dual mandate means little in practice. Monetary dominance is (a) reflected in the current conduct of monetary policy and (b) puts the price stability mandate first. There is no analysis of the political justifications for this.
\end{itemize}

\subsubsection{The technocratic legitimation claim}

\begin{itemize}
    \item \textbf{Fischer (1995)} --- Instrument independence justified by technical complexity. Monetary policy is too complex for democratic deliberation; it requires expert insulation.

    \item \textbf{Does double work} --- Justifies CBI on meritocratic grounds AND pre-emptively delegitimises democratic oversight as incompetent.

    \item \textbf{Flag} --- Bernanke in Section 4.3 confirms the informational asymmetry this claim relies on. Posen/Cochrane (2025) essentially restate this claim as well.
\end{itemize}

% -------------------------------------------------------
\subsection{The Political Science Literature on CBI (\textasciitilde400 words)}

\textit{Establish two core facts: (1) There is a non-engaged-with political science CBI literature, particularly regarding the democratic implications of monetary policy. (2) Central banks have a unique relationship with finance.}

\begin{itemize}
    \item \textbf{Literature on CBI and non-neutrality of central banks/bankers}
    \begin{itemize}
        \item \textbf{Adolph (2018)} --- Shadow principals: career incentives shape central banker preferences before and after office. Greater legal independence makes central bankers more powerful agents and more valuable to shadow principals. Critiques the ``myth of neutrality.''

        \item \textbf{Jacobs and King (2018)} --- The Fed has historically justified growing its powers through technocratic merit.

        \item \textbf{Zingales (2013)} --- Economists face the same capture mechanisms as regulators: career incentives, funding dependencies, professional socialisation. Pro-market findings circulate more widely and define what is acceptable to conclude.
    \end{itemize}

    \item \textbf{Downey (2025)} --- Double delegation of monetary powers.

    \item \textbf{Federal Advisory Council finding} --- Only the four largest banks on the FAC demonstrably influence the Federal Funds rate. The closest thing to direct empirical evidence of monetary policy capture.

    \item \textbf{Mabbett and Schelkle} --- Broader discussion of the Fed's ``loneliness.''

    \item \textbf{The missing interaction between literatures} --- Economics and political science study CBI without citing each other. The economics community is institutionally embedded; the political science critique has no equivalent footprint. This non-interaction is itself evidence.
\end{itemize}

% -------------------------------------------------------
\subsection{Regulatory Capture (\textasciitilde500 words)}

\begin{itemize}
    \item \textbf{Dal B\'{o} (2006): broad definition} --- Capture includes monetary policy explicitly. Pre-empts the objection that this only applies to utility regulation.

    \item \textbf{Stigler-Peltzman black box} --- Stigler: regulation is acquired by the industry and designed primarily for its benefit. Peltzman extends this into a model. Neither explains the internal mechanism --- Dal B\'{o} opens the box.

    \item \textbf{Kwak (2013): cultural capture} --- Regulators influenced through group identification and socialisation, not bribes. No explicit coordination required.

    \item \textbf{Instrumental vs.\ structural distinction} --- Instrumental requires observable coordination: lobbying, revolving doors, bribes. Structural operates through institutional design that aligns interests before influence is exercised.

    \item \textbf{GAO reports (2017, 2019)} --- Official acknowledgement of capture risk at the Fed and FDIC.

    \item \textbf{Posen (1995)} --- ``National differences in the degree of central bank independence in the postwar period resulted instead from differences in financial sector opposition to inflation'' (p.\ 253). Reveals the political coalition sustaining CBI.

    \item \textbf{Epistemic capture as extension} --- The captured community is the one that got institutionalised. The non-interaction between the two literatures is not accidental --- it is structural.
\end{itemize}

% -------------------------------------------------------
\section{Methodology (\textasciitilde950 words)}

\subsection{Research Design (\textasciitilde300 words)}

\begin{itemize}
    \item \textbf{Theory-testing single case study: the Federal Reserve} --- The Fed is the appropriate case because of its outsized importance in the development of CBI globally and the economics literature more broadly. The dual mandate is directly referenced in the index critique.

    \item \textbf{Process tracing (Beach and Pedersen 2019)} --- ``A research method for tracing causal mechanisms using detailed, within-case empirical analysis of how a causal mechanism operated in real-world cases'' (p.\ 1); ``we shift the analytical focus from causes and outcomes to the hypothesized causal mechanisms in between'' (p.\ 1). Inferences are made using ``the correspondence between the hypothetical empirical fingerprints that might have been left by the operation of a mechanism and the actual empirical observations we find in a case'' (p.\ 3).
    \begin{itemize}
        \item \textbf{Bennett and Checkel (2015)} --- Complementary guidelines for applying process tracing in political science, particularly on using within-case evidence to evaluate competing causal explanations.
    \end{itemize}

    \item \textbf{A sequential causal chain} --- Three sections are causally linked legs of one mechanism: the epistemological prior (4.1) produces the assumptions; those assumptions travel through personnel and institutional discourse (4.2); that transmission forecloses legislative revision (4.3). Each step is downstream of the prior one.
\end{itemize}

\subsection{Causal Chain and Evidentiary Tests (\textasciitilde400 words)}

\begin{itemize}
    \item \textbf{Step 1 --- hoop test: epistemological assumptions encode financial preferences} --- Necessary condition and origin of the chain. If the literature's assumptions don't systematically favour creditor interests, the whole mechanism fails at its source.

    \item \textbf{Step 2 --- hoop test: assumptions transmitted through institutional vessels} --- Necessary condition and middle leg. The assumptions from Step 1 must travel from academic literature into institutional practice through specific, traceable channels.

    \item \textbf{Step 3 --- smoking gun: legislative revision foreclosed using the same assumptions} --- The chain closes here. Legislative inaction defended by the same assumptions that originated in Step 1 is the circularity that confirms the mechanism is operating end-to-end.

    \item \textbf{Hoop test (Van Evera 1997)} --- Necessary but not sufficient: absent fingerprint kills the hypothesis; present fingerprint keeps it alive but doesn't confirm it.

    \item \textbf{Smoking gun test (Van Evera 1997)} --- Sufficient but not necessary: if found, strongly confirms the hypothesis; its absence doesn't eliminate it. Decisive because the evidence would be very hard to explain under any rival hypothesis.

    \item \textbf{Rival hypothesis limitation} --- \textit{[TO REVISIT]} Technocratic legitimacy (Posen) can explain Step 3 in isolation but can't explain why the measurement apparatus in Step 1 encodes financial preferences, or why the personnel patterns in Step 2 reproduce them.
\end{itemize}

\subsection{Rival Hypothesis (\textasciitilde250 words)}

\begin{itemize}
    \item \textbf{Posen (PIIE 2025): technocratic legitimacy} --- CBI works; legislators defer because it is rational to defer given the Fed's track record.

    \item \textbf{Insufficient} --- Explains the outcome but not the assumptions, not the personnel patterns, not the redirected reform energy.

    \item \textbf{The rival hypothesis is itself a product of the epistemic environment under scrutiny} --- Posen's argument reproduces the assumptions of the pro-CBI literature. Its inability to see capture is exactly what the capture thesis predicts.
\end{itemize}

% -------------------------------------------------------
\section{Analysis (\textasciitilde5,750 words)}

\subsection{The Epistemological Prior: Compromised Starting Conditions (\textasciitilde1,800 words)}

\textit{This is the load-bearing section. The argument is not about people protecting their money or career incentives --- it is that there is an entire literature ingraining special interests at the very beginning of what presents itself as non-partisan analysis. Even if policymakers and bureaucrats were to act and follow the literature strictly, they would reach the same conclusions about the merits of CBI. The question this section asks is how these two literatures --- economics and political science --- came to such radically different conclusions, and how the conclusions of one came to be encoded in institutional design while the other did not.}

\begin{itemize}
    \item \textbf{Price stability as unexamined social optimum} --- Kydland-Prescott assume low inflation is consistent with the public's preferences. They never defend that claim. Every subsequent paper inherits it.

    \item \textbf{Rogoff's third-best treated as first-best} --- The empirical literature takes the conservative banker solution and treats it as the unambiguous social optimum. Rogoff's own ranking, his employment variance caveat, his finite inflation aversion condition --- all dropped.

    \item \textbf{First-best encodes neoclassical labour market assumptions} --- Rogoff's preferred solution assumes market employment is being suppressed by distortions like unemployment insurance. That is a contestable normative claim, not a neutral starting point.

    \item \textbf{The free lunch claim} --- Alesina and Summers say lower inflation, no output cost. Their own caveat (``not conclusive'') disappears from the consensus. The trade-off Rogoff acknowledged is gone.

    \item \textbf{Technocratic claim encoded in index literature} --- The indices are presented as technical instruments. They encode normative preferences about what monetary policy should do. Expertise insulates the measurement apparatus from critique.

    \item \textbf{De jure measurement conflates mandate content with operational autonomy} --- The indices measure what the mandate contains, not how independently it is pursued. These are categorically different things.

    \item \textbf{Powell example: mandate revision toward employment lowers Romelli score} --- If Congress exercised its constitutional authority to revise the mandate, the Fed's independence score would fall. That proves the circularity. What is being measured is not independence --- it is the content of the mandate.

    \item \textbf{US/Turkey anomaly} --- The US scores below Turkey on de jure indices. Turkey is Posen's own example of a country that lost independence and got inflation. The qualitative and quantitative evidence point in opposite directions inside the same literature.

    \item \textbf{Two non-interacting communities demonstrated textually} --- Romelli doesn't cite Dal B\'{o}. PIIE contributors don't engage the capture literature. The non-interaction can be documented, not just asserted.

    \item \textbf{The incommensurability of starting questions} --- The economics literature asks how to optimise monetary policy. The political science literature asks whose interests monetary policy serves. These are not competing answers to the same question --- they are different questions entirely.

    \item \textbf{Schneider, Strahan and Yang (2023)} --- Capture framework applied to Fed stress testing in a peer-reviewed journal. The argument is established in the literature even before applying it to monetary policy.

    \item \textbf{Conclusion} --- Following the economics literature strictly would lead to pro-CBI conclusions. This leaves no room for political critique.
\end{itemize}

% -------------------------------------------------------
\subsection{Vessels of Transmission (\textasciitilde1,600 words)}

\textit{The prior doesn't stay in journals. It travels into institutions through specific channels --- and those channels are what this section traces.}

\subsubsection{How the Prior Travels: Governance, Communication, and Expertise (\textasciitilde800 words)}

\begin{itemize}
    \item \textbf{Downey: Fed never fully public} --- Reserve bank boards are two-thirds banker or banker-appointees. The primary dealer system concentrates monetary transmission in a small set of major financial institutions. Financial interests are designed into the institution before capture is even needed.

    \item \textbf{Bernanke (\textit{The Courage to Act}, 2015)} --- Congress didn't understand what QE was while it was happening. The Fed's own chair confirms Dal B\'{o}'s asymmetric information mechanism. The technocratic claim isn't just a justification --- it actively disables oversight.

    \item \textbf{Schnabel (2026)} --- ``The pandemic episode was, hence, a testament to monetary dominance. It underscored the strength of an institutional framework anchored in the primacy of price stability.'' Reproduces the epistemological prior in real time. Her financial dominance framing mirrors Lindblom's market as prison argument from the other direction --- both describe financial sector structural power constraining policy, but Schnabel presents it as a condition to manage rather than a form of capture.

    \item \textbf{PIIE 2025 as institutional exhibit} --- The monetary policy mandate is treated as beyond democratic contestation. Capture risk is acknowledged, but only for supervisory functions. The prior is being reproduced in real time.

    \item \textbf{Adolph: greater legal independence amplifies shadow principal influence} --- More independent central bank = more powerful agent = more valuable to financial sector principals.

    \item \textbf{Rockefeller/Wriston historical asymmetry} --- Financial sector critique of the Fed has always targeted specific constraints, not independence itself. After Volcker, the critique flips to attacking looseness (Druckenmiller). The asymmetry runs historically deep --- it is structural, not incidental.
\end{itemize}

\subsubsection{Revolving Door (\textasciitilde800 words)}

\begin{itemize}
    \item \textbf{The goal of this section} --- Show that the prior reproduces itself through personnel. The career pipeline from finance to central bank to finance ensures the epistemic community that produces the literature also staffs the institution.

    \item \textbf{ABA revolving door: Olson; Duke} --- The same individuals move between the ABA and the Fed Board of Governors. The epistemic community reproduces itself through appointments.

    \item \textbf{Bowman: community bank $\rightarrow$ Vice Chair for Supervision $\rightarrow$ SVB reframing} --- Adolph's shadow principals mechanism made concrete. Career trajectory predicts policy outcomes. SVB framed as a management failure, not a structural regulatory failure. Supervision staff cut 30\%. Post-SVB reforms reversed.

    \item \textbf{Wirsching (2018)} --- Finance-sector backgrounds empirically predict deregulatory outcomes. Pattern predicted by Adolph, explained by Kwak.

    \item \textbf{Federal Advisory Council} --- The four largest banks have a formal governance channel that demonstrably influences the Federal Funds rate. This isn't revolving door dynamics --- it is structural institutional design.

    \item \textbf{Kwak: cultural capture through socialisation} --- Regulators identify with regulated firms through repeated interaction. No coordination required. The prior reproduces through professional culture.
\end{itemize}

% -------------------------------------------------------
\subsection{Legislative Blockage (\textasciitilde2,350 words)}

\textit{The defences of legislative inaction use the same assumptions exposed in Section 4.1. That's the circularity: the literature that encodes financial preferences is the same literature mobilised to argue revision is unnecessary.}

\begin{itemize}
    \item \textbf{Volcker 1979: the original encoding event} --- Full regime shift toward price stability with no legal change. The dual mandate was operationally overwritten without democratic revision. Structural capture came before CBI entrenchment, not after.

    \item \textbf{Lindblom (1982): market as prison} --- Any attempt to revise the mandate triggers financial market volatility. No lobbying required. The structural deterrent grows as the financial sector grows --- UK banking sector went from roughly 100\% to 450\% of GDP since the 1970s. Schnabel's financial dominance framing confirms this mechanism from within the central banking literature itself.

    \item \textbf{Dodd-Frank as failed critical juncture} --- Post-2008 was the strongest reform moment in decades. Reform got channelled into adjustments --- Class A director reform, stress testing --- that foreclosed fundamental revision. QE had already recapitalised the financial sector, increasing its political power precisely when reform pressure was greatest (Jacobs and King 2018 --- ``political multiplier effect'').

    \item \textbf{ABA/BPI stress test litigation} --- Challenges the Fed when it imposes constraints; defends it against democratic oversight. That asymmetry is the behavioural signature of structural alignment.

    \item \textbf{McCubbins and Schwartz (1984): fire alarm oversight blocked} --- Congress only intervenes when sectoral interests trigger the alarm. The financial sector makes sure the alarm doesn't get pulled on mandate questions.

    \item \textbf{Bernanke/asymmetric information} --- Congress can't revise what it doesn't understand. The technocratic claim does active legislative work --- it doesn't just justify independence, it disables the mechanism that would produce revision.

    \item \textbf{PIIE 2025 as epistemic closure exhibit} --- The expert community advising Congress on monetary policy shares the same assumptions exposed in Section 4.1. The defences of independence are the prior in institutional form.

    \item \textbf{The circularity stated explicitly} --- The literature encoding financial preferences is the same literature mobilised to defend against legislative revision. The rival hypothesis isn't an independent check --- it's produced by the same epistemic environment.

    \item \textbf{Rival hypothesis engagement} --- Technocratic legitimacy explains the outcome. It can't explain the measurement bias, the personnel patterns, or the redirected reform energy. It needs three independent coincidences. Epistemic capture explains all three with one mechanism.
\end{itemize}

% -------------------------------------------------------
\section{Conclusion (\textasciitilde650 words)}

\begin{itemize}
    \item \textbf{Epistemic capture as unifying mechanism} --- The prior gets produced in a compromised epistemic environment $\rightarrow$ travels through personnel and institutional discourse $\rightarrow$ makes democratic revision structurally unthinkable. The cycle is self-reinforcing.

    \item \textbf{The ``ingrained'' framing} --- Financial interests are not external to the Fed's legitimating framework. They were encoded in it from the beginning.

    \item \textbf{Contribution 1} --- Epistemic capture as a mechanism; extends Dal B\'{o}/Kwak capture typology to monetary policy.

    \item \textbf{Contribution 2} --- CBI measurement is circular. The literature can't neutrally evaluate the institution it legitimates because it was shaped by the same interests the institution serves.

    \item \textbf{Technocratic claim as mechanism of circularity} --- Expertise justifies insulation $\rightarrow$ insulation means assumptions go unchallenged $\rightarrow$ unchallenged assumptions validate the expertise. The loop closes.

    \item \textbf{Implication} --- The policy consensus defending CBI rests on a literature whose starting conditions were compromised. It can't be taken as independent validation.

    \item \textbf{Caveat} --- This is not an argument for Trump's ability to remove Powell. That's a different problem --- authoritarianism and central banking --- and a separate research agenda. This argument is about democratic design, particularly on the ``goal'' side of CBI.

    \item \textbf{Closing} --- The question isn't whether the Fed should be independent from political interference. It's an analysis of the epistemological assumptions that underpin that independence. It's step 1 of building a more inclusive financial system.
\end{itemize}

\vspace{1cm}
\noindent\textit{Total: \textasciitilde10,000 words}

\end{doublespace}

\end{document}
