\documentclass[12pt,a4paper]{article}

% Encoding
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% Page layout
\usepackage{geometry}
\geometry{a4paper,left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm}

% Font - Arial (LSE requirement)
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}

% Spacing
\usepackage{setspace}

% Bibliography
\usepackage[
backend=biber,
style=apa,
url = false,
doi = false,
isbn=false,
]{biblatex}
\addbibresource{bibliography.bib} %Imports bibliography file
\usepackage[hidelinks]{hyperref}

% Headers - candidate number (LSE requirement)
\usepackage{fancyhdr}
\setlength{\headheight}{14.5pt}
\pagestyle{fancy}
\fancyhf{}
%\rhead{\small Candidate Number: XXXXX}
\rfoot{\thepage}
\renewcommand{\headrulewidth}{0pt} %Links in-text citations to references

%Fix hbox warning
\sloppy

% Misc
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{appendix}

\begin{document}

\begin{doublespace}
% Cover page - Moodle requirement
\thispagestyle{empty}
\pagenumbering{gobble}                          % suppress page numbers
\begin{center}
    \vspace*{2cm}
    {\large GV499 Dissertation\\
    MSc Public Policy and Administration \\
    Department of Government\\
    London School of Economics}\\
    \vspace{2cm}
    {\Large\textbf{Captured by Design: Central Bank Independence as Structural Regulatory Capture}}\\
    \vspace{2cm}
    Candidate Number: XXXXX\\
    \vspace{0.5cm}
    Word Count: XXXXX\\
    \vspace{0.5cm}
    \today
\end{center}
\newpage

% Abstract
\section*{Abstract}
\vspace{\baselineskip}

My abstract goes here

\newpage



\tableofcontents
\newpage

% Double spacing - Moodle requirement

\pagenumbering{arabic}                          % page numbers start here
\setcounter{page}{1}
%\input{sections/introduction}
%\input{sections/literature}
%\input{sections/theory}
%\input{sections/empirical}
%\input{sections/conclusion}

%If you getg sick of the compiling extra files on Positron,
%motify settings.json in positron:
%    "latex-workshop.latex.autoClean.run": "onBuilt",
 %   "latex-workshop.latex.clean.method": "glob",
 %   "latex-workshop.latex.clean.fileTypes": [
 %       "*.aux", "*.bcf", "*.bbl", "*.blg", "*.idx", "*.ind", "*.lof", "*.lot", 
 %       "*.out", "*.run.xml", "*.toc", "*.acn", "*.acr", "*.alg", "*.glg", "*.glo", 
 %       "*.gls", "*.ist", "*.fls", "*.log", "*.fdb_latexmk", "*.synctex.gz", "*.synctex"
 %   ]



\newpage
\section{Introduction}

The interests of financial institutions were encoded in the institutional structures governing the Federal Reserve. These interests can be traced from the epistemological assumptions rooted in economics literature on central bank independence, through personnel and discourse, into consistent blockage of legislative revision of the Fed's dual mandate. The end result is a system that reflects regulatory capture without requiring explicit coordination or lobbying among firms; private interests instead only have to promote inaction.


\section{Background and Theory}

There is little cross-discipline analysis on central bank independence. While the economics literature generally promotes findings regarding the merits of CBI, political science literature is more critical, arguing about the distributional consequences of monetary policy and the implications of CBI on democratic institutions. In this paper, I apply the findings of political science literature to economics literature.



\subsection{Economics Central Bank Independence Literature}

The theoretical foundation for central bank independence lies in a few seminal works.  \textcite{kydland_rules_1977} were the first to apply the time-inconsistency and credible commitment public policy problem to the inflation-unemployment tradeoff detailed in economics literature. In brief, in a model economy with rational agents and a government seeking to maximize some ``social objective function," governments will naturally have a temptation to implement a surprise expansionary (inflationary) policy that temporarily pushes down unemployment \parencite[473]{kydland_rules_1977}. Rational agents predict this action, so they set their inflation expectations higher, meaning that the surprise action the government would implement would trigger even higher inflation. This principle makes credible commitment by governments challenging, so the end result is heightened inflation expectations from agents. Inflation becomes anchored at a level that is higher than what the social objective function deems optimal.

%This is analagous to agents building homes in a flood-plane, despite government assurances that they would not receive flood protection. Once the homes are built, the government will have the incentive to reverse course and protect the homes, so agents predict this policy-reversal and build their homes anyway \parencite[p. 477]{kydland_rules_1977}. 

\textcite{barro_rules_1983} apply this logic directly to monetary authorities. First, note that they (along with much other theoretical literature on monetary policy) assume that monetary policy is neutral in the long-run. This principally means that while the amount of currency in circulation may increase with expansionary monetary policy, the relative price between goods should, in theory, not change after a change in monetary policy \parencite[49-50]{downey_economic_2025}. Given this, \textcite{barro_rules_1983} suggests that monetary policy can only reap benefits if there is a short-run surprise, before prices can readjust. However, rational agents incorporate this in their inflation expectations, pushing up inflation without reaping any rewards if monetary policy is conducted under discretion. While a central bank may seek to preserve its reputation, thereby creating some credibility regarding its policies, this is only a partial substitute. \textcite{barro_rules_1983} argue that the optimal setup is a central bank that is entirely constrained by rules, thereby establishing credibility and anchoring inflation expectations of agents. \textcite[1177]{rogoff_optimal_1985} provide several possible solutions, inclucing removing appointing a ``conservative central banker" who is ``known to place a greater weight on inflation stabilization" than the social objective function would suggest. They will be less tempted to implement inflationary policy, and rational agents set inflation expectations lower as a consequence. While they acknowledge that this may not be theoretical optimal, the conservative central banker argument underpins much of the empirical literature defending CBI.


Empirical analysis on the merits of central bank independence often starts with references to \textcite{grilli_political_1991}, \textcite{cukierman_measuring_1992}, and \textcite{alesina_central_1993}. The former of these two papers take observable characteristics of the institutional setup of central banks, compile these observable metrics into an index that measures the degree of independence that a central bank has. \textcite{alesina_central_1993} then applied various ind, and then produce a statistical model measuring how different levels of CBI influence economic variables like output and inflation. This general approach has been repeated and adjusted in the literature, most recently, by \textcite{romelli_political_2022} and \textcite{adrian_new_2024}. Arguing for the merits of a given index or strategy is beyond the scope of this paper, but they all share a basic theoretical basis in the arguments initially presented by \textcite{kydland_rules_1977}, \textcite{barro_rules_1983}, and \textcite{rogoff_optimal_1985}. I will present additional analysis on these empirics in section XXXX. However, these empirical papers generally find one result, namely that the politicization of monetary policy rases inflation without boosting economic growth \parencite[4]{cochrane_central_2025}. 

Crucially, the literature distinguishes between operational independence and goal independence, and this is important for the Fed. While political institutions should determine the goals of monetary policy (US Congress for the goals of the Federal Reserve), the Fed must remain operationally independent in achieving those goals \parencite[3]{cochrane_central_2025, debelle_how_1994}. Today, the Federal Reserve his a tripartite mandate of  promoting the goals of ``maximum employment, stable prices, and moderate long-term interest rates," set about the a 1977 amendment to the Federal Reserve Act passed by Congress. The Federal Reserve coloquially refers to this as the ``dual mandate" of stable prices and maximum employment, dropping the moderate long-term interest rate component because of its ``redundancy" \parencite[15]{downey_economic_2025}. 

\subsection{Political Science Literature on CBI}

%``In other words, no one is advocating that the Fed should be able to do whatever it wants to do, and no one serious ever did. Elected officials, Congress in the US system, must set the Fed’s policy goals" 

\subsection{Regulatory Capture}

Conceptualizations of regulatory capture are often rooted in the seminal work by \cite{stigler_theory_1971}. 


\section{Methodology}

\subsection{Research Design}

\subsection{Causal chain and evidentiary tests}

\subsection{Rival hypothesis}

\section{Analysis}

\subsection{Empirical Evidence and CBI Indices}

As previously mentioned, \textcite{grilli_political_1991}, \textcite{cukierman_measuring_1992}, and \textcite{alesina_central_1993} are all widely cited as providing the core empirical papers on the merits of CBI. THIS PARAGRAPH NEEDS TO BE FIXED. SEE THAT CHATS.
%However, it is worth noting that these papers are not complete in their analysis, and they all recognize their own limitations. In particular, measuring the operational independence that exists between a central bank in its decisions and political authorities is intrinsically different to measure. The human relationships between these figures, and the ``true" justifications why certain decisions were made, are not easy to measure. Influence is frequently tied to actions outside public knowledge, and it is impossible to know the full extent of influence.

To get around this, these papers focus on measures of ``de jure" central bank independence. They build indices based on observable shifts in laws governing monetary policy over time. The advantage of this approach is the grounding in a tangible, measurable action that can be easily and clearly translated into data. However, the role that laws play in changing the ``real" influence of a bank can be questionable. The stability of the legal institutions that enforce these laws greatly influence the practical influence of the laws over CBI, yet this is difficult to incorporate. At this point, though, within the academic literature, this is the settled on as the best empirical, data-driven methodlogy for measuring central bank independence.


The most recent, peer-reviewed publicly available CBI indices from \textcite{romelli_trends_2024} and \textcite{garriga_revisiting_2025}.

Before analyzing the construction of these indices, it's



\begin{table}[htbp]
\centering
\caption{Central Bank Independence Scores: United States vs.\ Turkey (2023)}
\label{tab:us_turkey_cbi}
\begin{tabular}{llccc}
\toprule
\textbf{Index / Sub-index} & \textbf{Source} & \textbf{USA} & \textbf{Turkey} & \textbf{Gap} \\
 & & \textbf{(2023)} & \textbf{(2023)} & \textbf{(Turkey $-$ USA)} \\
\midrule
\multicolumn{5}{l}{\textit{CBIE Index (Romelli 2022, 2024)}} \\
\quad Overall & Romelli (2022, 2024) & 0.614 & 0.801 & +0.187 \\
\quad Board independence & Romelli (2022, 2024) & 0.506 & 0.590 & +0.084 \\
\quad Policy independence & Romelli (2022, 2024) & 0.700 & 0.800 & +0.100 \\
\quad \textbf{Objectives} & Romelli (2022, 2024) & \textbf{0.250} & \textbf{1.000} & \textbf{+0.750} \\
\quad Lending restrictions & Romelli (2022, 2024) & 0.938 & 1.000 & +0.063 \\
\quad Financial independence & Romelli (2022, 2024) & 0.667 & 0.667 & \phantom{+}0.000 \\
\quad Reporting \& accountability & Romelli (2022, 2024) & 0.625 & 0.750 & +0.125 \\
\midrule
\multicolumn{5}{l}{\textit{Comparison indices}} \\
\quad GMT Index (normalised) & Grilli et al.\ (1991) & 0.750 & 0.813 & +0.063 \\
\quad CWN Index (unweighted) & Cukierman et al.\ (1992) & 0.808 & 0.909 & +0.101 \\
\quad CWN Index (weighted) & Cukierman et al.\ (1992) & 0.735 & 0.867 & +0.132 \\
\bottomrule
\end{tabular}
\smallskip
\begin{minipage}{\linewidth}
\footnotesize\textit{Notes:} All scores are on a $[0,1]$ scale where 1 denotes maximum independence. Data sourced directly from \textcite{romelli_political_2022} and \textcite{romelli_trends_2024}. The Objectives sub-index measures the degree to which price stability is the sole or primary statutory mandate; the Federal Reserve's dual mandate (price stability and maximum employment) scores 0.25. The bolded Objectives row drives the majority of the overall gap between the two countries.
\end{minipage}
\end{table}

\begin{table}[htbp]
\centering
\caption{Central Bank Independence Scores: United States vs.\ Turkey, ]tex== Dataset (2023)}
\label{tab:garriga_us_turkey}
\begin{tabular}{lccc}
\toprule
\textbf{Index / Sub-index} & \textbf{USA} & \textbf{Turkey} & \textbf{Gap} \\
 & & \textbf{(2023)} & \textbf{(2023)} & \textbf{(Turkey $-$ USA)} \\
\midrule
\multicolumn{5}{l}{\textit{CWN Index (Garriga 2025)}} \\
\quad Overall (unweighted) & 0.391 & 0.861 & +0.470 \\
\quad Overall (weighted)   & 0.475 & 0.899 & +0.424 \\
\midrule
\multicolumn{5}{l}{\textit{Sub-indices}} \\
\quad CEO independence        & 0.438 & 0.645 & +0.207 \\
\quad \textbf{Objectives}    & \textbf{0.400} & \textbf{0.800} & \textbf{+0.400} \\
\quad \textbf{Policy formulation} & \textbf{0.100} & \textbf{1.000} & \textbf{+0.900} \\
\quad Limitations on lending & 0.626 & 1.000 & +0.374 \\
\bottomrule
\end{tabular}
\smallskip
\begin{minipage}{\linewidth}
\footnotesize\textit{Notes:} All scores are on a $[0,1]$ scale where 1 denotes maximum independence. Data sourced from \textcite{garriga_revisiting_2025}, following \textcite{cukierman_measuring_1992} coding criteria. The Objectives sub-index measures the degree to which price stability is the sole or primary statutory mandate; the Federal Reserve's dual mandate (price stability and maximum employment) scores 0.40. The bolded rows drive the majority of the overall gap between the two countries.
\end{minipage}
\end{table}

I SHOULD TALK ABOUT THE PAPERS THAT CRITICIZE, IN PARTICULAR, ALESINA. 

\subsection{Epistemological Assumptions in the Pro-CBI Literature}

%Normative Loading
The first evidence for structural regulatory capture should be in the epistemic environment that CBI is built upon. In particular, there is a significant economics literature defending CBI. Many policymakers in central banking that this is a well-established consensus, and few mainstream policymakers openly criticize this consensus. If central bank independence reflects the interest of finance, then, principally, those interests should be reflected in assumptions that govern empirical and theoretical analysis of central banking. 


The presentation of the time inconsistency problem in \textcite[475]{kydland_rules_1977} begins with an ``agreed-upon social objective function" that is ``assumed to exist." Dissect page 479 and 480; there is a general push and pull 

Barro and gordon is really the place to begin, though. Let's start with monetary policy not mattering in output and unemployment in the long-run; only on inflation. But what are the costs of inflation? What if we sampled inflation on the basis of the poor? We have ``1" inflation rate, but really, we don't have one inflation rate, do we. We could mention the Fed's prefered PCE preference, food and gas, etc... To be fair, those aren't included right now anyway in the measure of inflation the monetary policy is crafted upon anyway.

Unemployment has a lasting benefit for those who get jobs.

On \textcite{barro_rules_1983},

the term $(a/2)(\pi_t)^2$ within their first equation $(a/2)(\pi_t)^2 - b_t (\pi_t - \pi_tTe$ measures the cost of inflation. Its worth questioning what hose exact costs are. To begin, \textcite[p. 104]{barro_rules_1983} note that  ``people generaIly regard inflation as very costly," but that ``economists have not presented very convincing arguments to explai .I these costs."

The paper starts with 


\textcite[357]{cukierman_measuring_1992}: ``"In Rogoff's (1985) terminology, the objectives variable measures the strength of the 'conservative bias' of the bank's charter." 

%Let's use the term normative loading.

Much of this literature is predicated on the idea that monetary policy is neutral in the long-run \parencite{downey_economic_2025}.

Normative loading on the preferences of the optimal objective are prevelant in both papers. In particular, there is a general assumption about what the optimal rate is and that it can be aggregated by the public. 

``The indifference curves imply that the socially preferred inflation rate is zero, which seems consistent with the public's preferences. We of course recognize that inflation is a tax on reserves and currency, and a more informed public might prefer some  positive or negative inflation rate. If so, x, need only be interpre as deviation from the optimal rate" \parencite[480]{kydland_rules_1977}.

They think ``of the monetary authority’s objective as reflecting the preferences of the “representative’ private agent. Ultimately, we express this objective as a function of actual and expected rates of inflation" \parencite[102]{barro_rules_1983}.


\subsection{Vessels of transmission}

\subsubsection{Governors' Speeches}


\subsubsection{Revolving Door}


\subsection{Legislative Blockage}


\section{Conclusion}



\end{doublespace}

% Bibliography - excluded from word count
\newpage

\printbibliography


\end{document}